---
title: "RandomForest on Differential Expression Proteins"
date: "Created: 2021-09-18 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introduction

1. Evaluating the important value of protein by recursive feature elimination

2. Adding protein into model when its spearman correlation value with any existing protein in the model was less than 0.7

3. Dividing data into 80% for trainset and 20% for testset

4. Calculating the performance using 10-fold cross-validation



Discovery Set: 1111

  * 485 HC
  
  * 145 AA

  * 481 CRC  
  
  
Validation Set: 476 

  * 202 HC
  
  * 62 AA

  * 212 CRC  

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(convert)
library(data.table)
library(caret)
library(pROC)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data
```{r}
ExprSet_Discovery <- readRDS("../../Result/Biomarker/RF_model/species_profile_Discovery.RDS")
ExprSet_Validation <- readRDS("../../Result/Biomarker/RF_model/species_profile_Validation.RDS")
CRC_HC_DEA <- fread("../../Result/Biomarker/RF_model/CRC_HC_DEA.csv")
```


### RandomForest model
```{r}
RF_model <- function(dataset=ExprSet_Discovery,
                     testres=CRC_HC_DEA,
                     group_info="SubGroup",
                     group_name=subgrp[c(1, 3)],
                     Pname="adj.P.Val",
                     Pval=0.05,
                     LogFC=0,
                     Multicol=TRUE,
                     Rho=0.7,
                     SplitPro=0.7,
                     kfold=5,
                     times=10,
                     oneFit=TRUE){
  
  # dataset=ExprSet_Discovery
  # testres=CRC_HC_DEA
  # group_info="SubGroup"
  # group_name=subgrp[c(1, 3)]
  # Pname="adj.P.Val"
  # Pval=0.05
  # LogFC=0
  # Multicol=FALSE
  # Rho=0.7
  # SplitPro=0.7
  # kfold=5
  # times=10
  # oneFit=TRUE

  # Differential Tax
  colnames(testres)[which(colnames(testres) == Pname)] <- "PValue"
  
  if(Pval == 0){
    dat_DEP <- testres %>% dplyr::select(FeatureID, PValue)
  }else{
    if(LogFC > 0){
      dat_DEP <- testres %>% dplyr::select(FeatureID, PValue, logFC) %>%
        filter(PValue < Pval) %>%
        filter(abs(logFC) > LogFC)    
    }else{
      dat_DEP <- testres %>% dplyr::select(FeatureID, PValue) %>%
        filter(PValue < Pval)    
    }  
  }
  print(dim(dat_DEP)[1])

  phenotype <- pData(dataset)
  profile <- exprs(dataset)
  colnames(phenotype)[which(colnames(phenotype) == group_info)] <- "Group"
  
  # dataset of Tax
  if(length(group_name) == 3){
    group_name2 <- c("CRC", "non_CRC")
    pheno <- phenotype %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[1], group_name2[1], group_name2[2])) %>%
      mutate(Group=factor(Group, levels = group_name2)) %>%
      column_to_rownames("SampleID")
    group_name <- group_name2
  }else{
    pheno <- phenotype %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name)) %>%
      column_to_rownames("SampleID")
  }
  print(table(pheno$Group))
  
  edata <- profile[rownames(profile)%in%dat_DEP$FeatureID, 
                   colnames(profile)%in%rownames(pheno)] %>% 
    data.frame()
  
  # Data of whole feature 
  mdat <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                       dplyr::select(SampleID, Group),
                     edata %>% t() %>% data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID")
  
  # remove multicollinearity
  if(Multicol){
    feature_cor <- Hmisc::rcorr(as.matrix(mdat[, -1]), type = "spearman")
    index_cor <- findCorrelation(feature_cor$r, cutoff = Rho, verbose = TRUE)
    feature_remain <- feature_cor$r[-index_cor, -index_cor]
    mdat_remain <- mdat %>% dplyr::select(c("Group", rownames(feature_remain)))     
  }else{
    mdat_remain <- mdat
  }
  
  # recursive feature elimination
  require(mlbench)
  rfeControl <- rfeControl(functions=rfFuncs, 
                           method="cv", 
                           number=10, 
                           rerank=TRUE)
  rfebreaks <- seq(2, ncol(mdat_remain), 2)
  rfefit <- rfe(mdat_remain[, -1],
                mdat_remain$Group,
                sizes=rfebreaks,
                rfeControl=rfeControl)
  
  rfeAccuracy <- tibble(Accuracy = signif(rfefit$results$Accuracy, 3),
                        Variables = rfefit$results$Variables) %>%
                 mutate(label=ifelse(Accuracy == max(Accuracy),
                           paste0("(Features=", Variables, ", Accuracy=", Accuracy,")"), NA))
  rfepl <- ggplot(data=rfeAccuracy, aes(x=Variables, y=Accuracy))+
            geom_point(color="grey", size=3, shape=19)+
            geom_line(color="black", linetype=1, size=1)+
            geom_text(aes(label=label), nudge_y=0.002)+
            annotate(geom="point",
                     x=rfeAccuracy[grep(max(rfeAccuracy$Accuracy), rfeAccuracy$Accuracy), ]$Variables,
                     y=max(rfeAccuracy$Accuracy), color="red", size=4)+
            labs(x="Features (Numbers)",
                 y="Accuracy (Bootstrap)")+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color='black'),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))
  
  # importance value 
  rfeImp <- varImp(rfefit) %>% 
    rownames_to_column("feature") %>%
    filter(feature%in%rfefit$optVariables) %>%
    arrange(desc(Overall))
  mdat_remain_v2 <- mdat_remain %>% dplyr::select(c("Group", rfeImp$feature))
  
  # Data Partition
  if(!is.null(SplitPro)){
    require(sampling)
    #set.seed(123)
    sample <- sampling::strata(mdat_remain_v2, 
                       stratanames = "Group", 
                       size = rev(round(as.numeric(table(mdat_remain_v2$Group)) * SplitPro)),
                       method = "srswor")
    trainData <- mdat_remain_v2[sample$ID_unit, ]
    testData <- mdat_remain_v2[-sample$ID_unit, ]
  }else{
    trainData <- mdat_remain_v2
    testData <- mdat_remain_v2
  }
  
  # normalization
  mdat_preProcess <- preProcess(mdat_remain_v2, method = c("center", "scale"))
  trainTransformed <- predict(mdat_preProcess, trainData)
  testTransformed <- predict(mdat_preProcess, testData)
  
  # dat_table <- as.matrix(trainTransformed %>% dplyr::select(-Group) %>% data.frame())
  # plot(density(dat_table))
  
  # construct model
  if(dim(trainTransformed)[1] > 10 * 8){
    kfold <- 10
  }else{
    kfold <- kfold
  }
  # N-repeat K-fold cross-validation
  myControl <- trainControl(method = "repeatedcv", 
                            number = kfold,
                            repeats = times,
                            search = "random",
                            summaryFunction = twoClassSummary,
                            classProbs = TRUE,                           
                            verboseIter = TRUE)
  
  if(oneFit){
    fit <- train(Group ~., 
                 data = trainTransformed,
                 method = "rf",
                 trControl = myControl,
                 tuneLength = 15,
                 metric = "ROC",
                 verbose = FALSE) 
    # Accuracy
    fit_Acc <- tibble(Feartures = fit$results$mtry,
                      ROC = signif(fit$results$ROC, 3),
                      Sens = signif(fit$results$Sens, 3),
                      Spec = signif(fit$results$Spec, 3)) %>%
                   mutate(label=ifelse(ROC == max(ROC), 
                              paste0("(Features=", Feartures, "\n ROC=", ROC,")"), NA))
    fit_final <- fit
    fit_Acc_final <- fit_Acc
    testData_final <- testTransformed    
    
  }else{
    # adding feature one by one according to their ranked importance value
    fit <- list()
    fit_Acc <- list()
    testData_list <- list()
    j <- 0
    for(i in 3:ncol(trainTransformed)){
      trainTransformed_cln <- trainTransformed[, c(1:i)]
      testTransformed_cln <- testTransformed[, c(1:i)]
      if(all(j == 0)){
        trainTransformed_cln_v2 <- trainTransformed_cln
        testTransformed_cln_v2 <- testTransformed_cln       
      }else{
        trainTransformed_cln_v2 <- trainTransformed_cln[, -j]
        testTransformed_cln_v2 <- testTransformed_cln[, -j]        
      }    
      train_cor <- Hmisc::rcorr(as.matrix(trainTransformed_cln_v2[, -1]), type = "spearman")
      train_index_cor <-  findCorrelation(train_cor$r, cutoff = Rho, verbose = TRUE)
      if(length(train_index_cor) == 0){
        # model 
        fit[[i-2]] <- train(Group ~., 
                      data = trainTransformed_cln_v2,
                      method = "rf",
                      trControl = myControl,
                      tuneLength = 15,
                      metric = "ROC",
                      verbose = FALSE)
        # Accuracy
        fit_Acc[[i-2]] <- tibble(Feartures = fit[[i-2]]$results$mtry,
                                 ROC = signif(fit[[i-2]]$results$ROC, 3),
                                 Sens = signif(fit[[i-2]]$results$Sens, 3),
                                 Spec = signif(fit[[i-2]]$results$Spec, 3)) %>%
                   mutate(label=ifelse(ROC == max(ROC), 
                              paste0("(Features=", Feartures, "\n ROC=", ROC,")"), NA)) %>%
                   mutate(Model_N=i-2)
        # testData
        testData_list[[i-2]] <- testTransformed_cln_v2
      }else{
        j <- c(j, i)
      }
    }
   
    # summary models
    # if(length(fit) > 1){
    #   print(summary(resamples(fit)))    
    # }
  
    # best model
    fit_Acc_All <- do.call(rbind, fit_Acc) %>%
      arrange(desc(ROC))
    
    fit_final <- fit[[fit_Acc_All$Model_N[1]]]
    fit_Acc_final <- fit_Acc[[fit_Acc_All$Model_N[1]]]
    testData_final <- testData_list[[fit_Acc_All$Model_N[1]]]
    
  }
  
  # the best Accuracy with optimal features
  fit_Acc_pl <- ggplot(data=fit_Acc_final, aes(x=Feartures, y=ROC))+
              geom_point(color="grey", size=3, shape=19)+
              geom_line(color="black", linetype=1, size=1)+
              geom_text(aes(label=label), nudge_y=-0.001)+
              annotate(geom="point",
                       x=fit_Acc_final[grep(max(fit_Acc_final$ROC),
                                                fit_Acc_final$ROC), ]$Feartures,
                       y=max(fit_Acc_final$ROC), color="red", size=4) +
              labs(x="Randomly Selected Predictors",
                   y="ROC (Repeated Cross-Validation)")+
              theme_bw()+
              theme(panel.background = element_rect(fill="transparent"),
                    axis.ticks.length = unit(0.4, "lines"),
                    axis.ticks = element_line(color='black'),
                    axis.line = element_line(size=.5, color="black"),
                    axis.title = element_text(color='black', size=12, face="bold"),
                    axis.text = element_text(color='black', size=10),
                    text = element_text(size=8, color="black", family="serif"))
    
  # evaluate the performance of model
  pred_raw <- predict(fit_final, newdata = testData_final, type = "raw")
  print(confusionMatrix(pred_raw, testData_final$Group))
  pred_prob <- predict(fit_final, newdata = testData_final, type = "prob")  
    
  # define roc and calculate AUC
  rocobj <- roc(testData_final$Group, pred_prob[, 1])
  auc <- round(auc(testData_final$Group, pred_prob[, 1]), 3)
  roc <- tibble(tpr=rocobj$sensitivities,
                  fpr=1 - rocobj$specificities)
  roc_pl <- ggplot(data=roc, aes(x=fpr, y=tpr))+
              geom_path(color="red", size=1)+
              geom_abline(intercept=0, slope=1, color="grey", size=1, linetype=2)+
              labs(x = "False Positive Rate (1 - Specificity)",
                   y = "True Positive Rate (Sensivity or Recall)")+
              annotate("text", x=.75, y=.25, label=paste("AUC =", auc),
                       size=5, family="serif")+
              coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
              theme_bw()+
              theme(panel.background = element_rect(fill="transparent"),
                    axis.ticks.length = unit(0.4, "lines"),
                    axis.ticks = element_line(color="black"),
                    axis.line = element_line(size=.5, color="black"),
                    axis.title = element_text(color='black', size=12, face="bold"),
                    axis.text = element_text(color='black', size=10),
                    text = element_text(size=8, color="black", family="serif"))
    
  # important features
  imp <- varImp(fit_final, scale=TRUE)$importance %>%
            setNames("Importance") %>%
            rownames_to_column("Features") %>%
            arrange(desc(Importance)) %>%
            dplyr::slice(1:max(fit_final$bestTune))
  imp$Features <- factor(imp$Features, levels = rev(as.character(imp$Features)))
  feature_pl <- ggplot(imp, aes(x=Features, y=Importance, fill=factor(Features))) +
      geom_bar(stat="identity", color="black")+
      xlab("")+
      guides(fill="none")+
      coord_flip()+
      theme_classic()+
      theme(panel.background = element_blank(),
            panel.grid = element_blank(),
            axis.line.x = element_line(size=.5, color="black"),
            axis.line.y = element_line(size=.5, color="black"),
            axis.ticks = element_line(color="black"),
            axis.title = element_text(color="black", size=12, face="bold"),
            axis.text.y = element_text(color="black", size=10, face="italic"),
            legend.position = "right",
            legend.background = element_blank(),
            legend.key = element_blank(),
            legend.text = element_text(size=8, face="italic"),
            text = element_text(size=8, color="black", family="serif"))
    
  # results
  res <- list(pl=list(rfe=rfepl, acc=fit_Acc_pl, roc=roc_pl, feature=feature_pl),
              model=list(rfe=rfefit, Repeatfit=fit, fit_acc=fit_Acc, fit=fit_final))
  return(res)
}

# validation
model_validation <- function(dataset=ExprSet_Validation,
                             datfit=CRC_HC_RF$model$fit,
                             group_name=subgrp[c(1, 3)]){
  
  # dataset=ExprSet_Validation
  # datfit=CRC_HC_RF$model$fit
  # group_name=subgrp[c(1, 3)]
  
  # dataset of DEP
  if(length(group_name) == 3){
    group_name2 <- c("CRC", "non_CRC")
    pheno <- pData(dataset) %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[1], group_name2[1], group_name2[2])) %>%
      mutate(Group=factor(Group, levels = group_name2)) %>%
      column_to_rownames("SampleID")
    group_name <- group_name2
  }else{
    pheno <- pData(dataset) %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name)) %>%
      column_to_rownames("SampleID")
  }
  
  profile <- exprs(dataset)
  edata <- profile[, colnames(profile)%in%rownames(pheno)] %>% 
    data.frame()
  
  # Data of whole feature 
  mdat <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                       dplyr::select(SampleID, Group),
                     edata %>% t() %>% data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID")
  
  # evaluate the performance of model
  pred_raw <- predict(datfit, newdata = mdat, type = "raw")
  print(confusionMatrix(pred_raw, mdat$Group))
  pred_prob <- predict(datfit, newdata = mdat, type = "prob")  
  
  # define roc and calculate AUC
  rocobj <- roc(mdat$Group, pred_prob[, 1])
  auc <- round(auc(mdat$Group, pred_prob[, 1]), 3)
  roc <- tibble(tpr=rocobj$sensitivities,
                fpr=1 - rocobj$specificities)
  roc_pl <- ggplot(data=roc, aes(x=fpr, y=tpr))+
            geom_path(color="red", size=1)+
            geom_abline(intercept=0, slope=1, color="grey", size=1, linetype=2)+
            labs(x = "False Positive Rate (1 - Specificity)",
                 y = "True Positive Rate (Sensivity or Recall)")+
            annotate("text", x=.75, y=.25, label=paste("AUC =", auc),
                     size=5, family="serif")+
            coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color="black"),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))
  
  # Accuracy
  CFM <- confusionMatrix(pred_raw, mdat$Group)
  CFM_index <- data.frame(Value=CFM$byClass) %>%
    rownames_to_column("Index") %>% 
    rbind(data.frame(Index="Accuracy", Value=as.numeric(CFM$overall[1])))
  
  res <- list(auc=auc, roc=roc_pl, index=CFM_index)
  
  return(res)
}

if(!dir.exists("../../Result/Biomarker/RF_model/")){
  dir.create("../../Result/Biomarker/RF_model/", recursive = T)
}
```


### 10-repeat 10-fold cross validation random forest
```{r}
# if(!file.exists("../../Result/Biomarker/RF_model/CRC_HC_RF_Discovery.RDS")){
#    CRC_HC_RF <- RF_model(
#                      dataset=ExprSet_Discovery,
#                      testres=CRC_HC_DEA,
#                      group_info="SubGroup",
#                      group_name=subgrp[c(1, 3)],
#                      Pname="adj.P.Val",
#                      Pval=0.05,
#                      LogFC=0,
#                      Multicol=FALSE,
#                      Rho=0.7,
#                      SplitPro=0.7,
#                      kfold=5,
#                      times=10,
#                      oneFit=TRUE)
#   saveRDS(CRC_HC_RF, "../../Result/Biomarker/RF_model/CRC_HC_RF_Discovery.RDS", compress = TRUE)
# }else{
#   CRC_HC_RF <- readRDS("../../Result/Biomarker/RF_model/CRC_HC_RF_Discovery.RDS")
# }


CRC_HC_RF <- RF_model(
                     dataset=ExprSet_Discovery,
                     testres=CRC_HC_DEA,
                     group_info="SubGroup",
                     group_name=subgrp[c(1, 3)],
                     Pname="adj.P.Val",
                     Pval=0.05,
                     LogFC=0,
                     Multicol=FALSE,
                     Rho=0.7,
                     SplitPro=0.7,
                     kfold=5,
                     times=10,
                     oneFit=TRUE)
CRC_HC_RF_validation <- model_validation(
                           dataset=ExprSet_Validation,
                           datfit=CRC_HC_RF$model$fit,
                           group_name=subgrp[c(1, 3)])

# while(with(CRC_HC_RF_validation, auc < 0.8 | index$Value[1] < 0.7 | index$Value[2] < 0.7))
while(with(CRC_HC_RF_validation, index$Value[12] < 0.7)){
  CRC_HC_RF <- RF_model(
                     dataset=ExprSet_Discovery,
                     testres=CRC_HC_DEA,
                     group_info="SubGroup",
                     group_name=subgrp[c(1, 3)],
                     Pname="adj.P.Val",
                     Pval=0.05,
                     LogFC=0,
                     Multicol=FALSE,
                     Rho=0.7,
                     SplitPro=0.7,
                     kfold=5,
                     times=10,
                     oneFit=TRUE)
  CRC_HC_RF_validation <- model_validation(
                             dataset=ExprSet_Validation,
                             datfit=CRC_HC_RF$model$fit,
                             group_name=subgrp[c(1, 3)])
}

saveRDS(CRC_HC_RF, "../../Result/Biomarker/RF_model/CRC_HC_RF_Discovery.RDS", compress = TRUE)
```


### systemic information
```{r}
sessionInfo()
```


### Reference

1. [Underdevelopment of the gut microbiota and bacteria species as non-invasive markers of prediction in children with autism spectrum disorder](https://pubmed.ncbi.nlm.nih.gov/34312160/)

2. [Recursive Feature Elimination](https://topepo.github.io/caret/recursive-feature-elimination.html)

3. [Summary multiple model's Accuracy](https://stackoverflow.com/questions/57939453/building-a-randomforest-with-caret)

