---
title: "Enterotype: Partition Around Medoids Clustering"
date: "Created: 2021-11-22 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introduction

Clustering fecal samples into enterotypes by using Partition Around Medoids Clustering(PAM) is common routine for subgroup population in metagenomic analysis. Here is the description of PAM.

> 1. Select k objects from the dataset as the initial k medoids.
>
> 2. If, some k objects are already marked to be used as medoids, use them as medoids in the data-set.
>
> 3. Calculate the dissimilarity matrix (Euclidean distance or Manhattan distance et al).
>
> 4. Assign other objects to their closest medoids.
> 
> 5. For each cluster, if you find any object that can reduce the average dissimilarity co-efficient, use it as that cluster’s medoid. Also, that object should be chosen as medoid which decreases the average dissimilarity co-efficient the most.
>
> 6. If you find new medoid, go back to step 4 and reiterate till you don’t get any new medoid.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(convert)
library(factoextra)
library(cluster)
library(clusterSim)
library(ade4)
library(ggplot2)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data 
```{r}
ExprSet <- readRDS("../../Result/Profile/species_profile.RDS")
```


### Function
```{r}
JSD_dist_Fun <- function(datset=ExprSet, 
                         pseudocount=0.000001, 
                         ...){
  
  # datset=ExprSet
  # pseudocount=0.000001
  
  
  KLD <- function(x, y){sum(x *log(x/y))}
  JSD <- function(x, y){sqrt(0.5 * KLD(x, (x+y)/2) + 0.5 * KLD(y, (x+y)/2))}

  inMatrix <- exprs(datset) %>% data.frame()
  matrixColSize <- length(colnames(inMatrix))
  matrixRowSize <- length(rownames(inMatrix))
  colnames <- colnames(inMatrix)
  resultsMatrix <- matrix(0, matrixColSize, matrixColSize)
  inMatrix <- apply(inMatrix, 1:2, function(x){ifelse(x==0, pseudocount, x)})
  
  for(i in 1:matrixColSize){
    for(j in 1:matrixColSize){ 
      resultsMatrix[i, j] <- JSD(as.vector(inMatrix[, i]), as.vector(inMatrix[, j]))
    }
  }
  colnames -> colnames(resultsMatrix) -> rownames(resultsMatrix)
  as.dist(resultsMatrix) -> resultsMatrix

  attr(resultsMatrix, "method") <- "dist"
  
  return(resultsMatrix)
}

PAM_Cluster_Fun <- function(x, k){
  
  # x=dat_JSD
  # k=best_k
  
  cluster <- as.vector(pam(as.dist(x), k, diss=TRUE)$clustering)
  
  return(cluster)
}


CH_Index_Fun <- function(datset=ExprSet,
                         pseudocount=0.000001,
                         distance.jsd=dat_JSD, 
                         kvector=c(1:20), 
                         clusterSim=TRUE){
  # datset=ExprSet
  # pseudocount=0.000001
  # distance.jsd=dat_JSD
  # kvector=c(1:20)
  # clusterSim=TRUE  
  
  inMatrix <- exprs(datset) %>% data.frame()
  dataframe <- apply(inMatrix, 1:2, function(x){ifelse(x==0, pseudocount, x)})

  if(clusterSim){
  	require(clusterSim)
  	kvector <- unique(as.integer(kvector))
  	nclusters <- NULL
		for(k in kvector){ 
			if(k==1){
				nclusters[k] <- 0
			}else{
				if(is.null(distance.jsd)){
				  stop("a distance matrix is needed")
				}else{
  				cluster <- PAM_Cluster_Fun(distance.jsd, k)
  				nclusters[k] <- index.G1(t(dataframe), cluster, d = distance.jsd, centrotypes = "medoids")				    
				}
			}
		}
	}else{
	  nclusters <- pamk(t(dataframe), criterion="ch", krange=kvector)$crit
	}
  
  return(nclusters)
}

if(!dir.exists("../../Result/Enterotype")){
  dir.create("../../Result/Enterotype", recursive = T)
}
```


### Run: JSD distance
```{r}
# calculate JSD distance
dat_JSD <- JSD_dist_Fun(datset = ExprSet)
# primary clustering 
dat_cluster <- PAM_Cluster_Fun(dat_JSD, k=4)
# determine best cluster
dat_nclusters <- CH_Index_Fun(datset = ExprSet, distance.jsd = dat_JSD)
plot(dat_nclusters, type="h", xlab="k clusters", ylab="CH index", main="Optimal number of clusters")

# final cluster result
k_best <- which(dat_nclusters == max(dat_nclusters), arr.ind = TRUE)
dat_cluster_final <- PAM_Cluster_Fun(dat_JSD, k = k_best)
mean(silhouette(dat_cluster_final, dat_JSD)[, k_best])
```

### Visualization: Principal Coordinates Analysis (PCoA)
```{r}
# s.class
obs.pcoa <- dudi.pco(dat_JSD, scannf = F, nf = 3)
s.class(obs.pcoa$li, 
        fac = as.factor(dat_cluster_final), 
        grid = F, 
        sub = "Principal coordiante analysis", 
        col=c(1, 2))

# ggplot2
PCoAFun <- function(dcluster=dat_cluster_final,
                    distance=dat_JSD){
  
  # dcluster=dat_cluster_final
  # distance=dat_JSD
  
  require(ape)
  require(vegan)
  pcoa <- pcoa(distance)
  eig <- pcoa$values[, "Eigenvalues"]
  eig_var <- eig[1:2]
  eig_var_explain <- round(eig_var/sum(eig), 4) * 100   
  # explains variable 
  explains <- paste0(paste0("PCoA", seq(2)), " (", paste0(eig_var_explain, "%"), ")")
  # principal component score of each sample
  score <- cbind(pcoa$vectors[, c(1:2)] %>% data.frame() %>%
                        setNames(paste0("PCoA", seq(2))), 
                      dcluster %>% data.frame() %>% 
                        setNames("Enterotype")) %>%
    mutate(Enterotype=paste0("E", Enterotype)) %>%
    mutate(Enterotype=factor(Enterotype))
  
  pl_scatter <- ggplot(score, aes(x=PCoA1, y=PCoA2))+
              geom_point(aes(fill=Enterotype), size=3, shape=21, stroke = .8, color = "black")+
              stat_ellipse(aes(color=Enterotype), level = 0.95, linetype = 1, size = 1.5)+
              labs(x=explains[1], y=explains[2])+
              scale_color_manual(values = c("red", "Blue"))+
              scale_fill_manual(values = c("red", "Blue"))+
              guides(color="none")+
              theme_classic()+
              theme(axis.title = element_text(size = 10, color = "black", face = "bold"),
                    axis.text = element_text(size = 9, color = "black"),
                    text = element_text(size = 8, color = "black", family = "serif"),
                    strip.text = element_text(size = 9, color = "black", face = "bold"), 
                    panel.grid = element_blank(),
                    legend.title = element_text(size = 11, color = "black", family = "serif"),
                    legend.text = element_text(size = 10, color = "black", family = "serif"),
                    legend.position = c(0, 0),
                    legend.justification = c(0, 0),
                    legend.background = element_rect(color = "black", fill = "white", linetype = 2, size = 0.5))
  
  group <- levels(score$Enterotype)
  cmp <- NULL
  for(i in 1:(length(group) -1 )){
    for(j in (i+1):length(group)){
      tmp <- c(group[i], group[j])
      if(is.null(cmp)){
        cmp[[1]] <- tmp
      }else{
        cmp[[i+1]] <- tmp
      }
    }
  }  
  
  pl_pcoa1_boxplot <- ggplot(score, aes(x=Enterotype, y=PCoA1, fill=Enterotype))+
        stat_boxplot(geom = "errorbar", width = .12)+
        geom_boxplot(width = .3, outlier.shape = 3, outlier.size = 1)+
        stat_summary(fun = mean, geom = "point", shape = 16, 
                     size = 2, color = "black")+
        guides(fill=F)+
        ggpubr::stat_compare_means(comparisons = cmp,
                           method = "wilcox.test", label = "p.label")+
        labs(x="", y="")+
        theme_bw()+
        scale_fill_manual(values = c("red", "Blue"))+
        coord_flip()+
        theme(axis.text = element_text(size = 9, color = "black"),
              axis.text.x = element_blank(),
              axis.ticks.x = element_blank(), 
              panel.grid = element_blank())
  
  pl_pcoa2_boxplot <- ggplot(score, aes(x=Enterotype, y=PCoA2, fill=Enterotype))+
        stat_boxplot(geom = "errorbar", width = .12)+
        geom_boxplot(width = .3, outlier.shape = 3, outlier.size = 1)+
        stat_summary(fun = mean, geom = "point", shape = 16, 
                     size = 2, color = "black")+
        guides(fill=F)+
        ggpubr::stat_compare_means(comparisons = cmp,
                           method = "wilcox.test", label = "p.label")+
        labs(x="", y="")+
        theme_bw()+
        scale_fill_manual(values = c("red", "Blue"))+
        theme(axis.text = element_text(size = 9, color = "black"),
              axis.text.y = element_blank(),
              axis.ticks.y = element_blank(), 
              panel.grid = element_blank()) 
  
  require(cowplot)
  pl <- plot_grid(plot_grid(pl_pcoa1_boxplot, pl_scatter, ncol=1, align = "hv", rel_heights = c(.5, 2)),
                  pl_pcoa2_boxplot, nrow=1, align = "hv", rel_widths = c(2, .5))
  
  return(pl)
}

PCoAFun(dcluster=dat_cluster_final, distance=dat_JSD)
```



### systemic information
```{r}
sessionInfo()
```


### Reference 

1. [BiotypeR](https://github.com/tapj/biotyper/blob/master/R/BiotypeR.r)

2 .[PAM Clustering Using R](https://mlforanalytics.com/2020/05/23/pam-clustering-using-r/)

3. [K-means vs K-mediods](https://www.datamining365.com/2020/03/partitional-clustering-k-means.html)

4. [R语言：肠型分析](https://www.jianshu.com/p/7d550681e37f)
