---
title: "MaAsLin2: Multivariable Association in Population-scale Meta-omics Studies"
date: "Created: 2021-10-09 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introduction

To further explore how facotrs impacted gut microbiome composition, investigating the correlations between individual factors and bacterial species via MaAsLin2 package

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
options(warn = 0)
library(dplyr)
library(tibble)
library(data.table)
library(Maaslin2)
library(ggplot2)
options(warn = 0)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data 
```{r}
ExprSet <- readRDS("../../Result/Profile/species_profile.RDS")
```


### Multivariable Association in Population-scale Meta-omics Studies
```{r}
Maaslin2Fun <- function(dataset=ExprSet,
                        group_info="SubGroup",
                        group_name=subgrp[c(1,3)],
                        occurrence=0.2,
                        taxa="species"){
  
  # dataset=ExprSet
  # group_info="SubGroup"
  # group_name=subgrp[c(1,3)]
  # occurrence=0.2
  # taxa="species"
  
  require(convert)
  phenotype <- pData(dataset)
  profile <- exprs(dataset)
  colnames(phenotype)[which(colnames(phenotype) == group_info)] <- "Group"

  pheno <- phenotype %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name))
  prof <- profile[, colnames(profile)%in%pheno$SampleID] %>% data.frame() %>% 
      rownames_to_column("tmp") %>%
    filter(apply(dplyr::select(., -one_of("tmp")), 1, function(x) {
            sum(x != 0)/length(x)}) > occurrence) %>%
    column_to_rownames("tmp") %>%
    t() %>% data.frame()
  
  mdat <- inner_join(pheno,
                     prof %>% rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID")
  
  input_data <- mdat[, c(ncol(pheno):ncol(mdat))] 
  input_metadata <- mdat[, c(1:(ncol(pheno)-1))]
  for(i in 1:ncol(input_metadata)){
    if(!is.numeric(input_metadata[, i])){
      input_metadata[, i] <- as.numeric(factor(input_metadata[, i]))
    }
  }
  
  outdir <- paste0("../../Result/MaAsLin2/", taxa)
  if(!dir.exists(outdir)){
    dir.create(outdir, recursive = TRUE)
  }
  
  # Run MaAsLine2
  fit_data <- Maaslin2(
                  input_data, 
                  input_metadata, 
                  outdir,
                  min_abundance = 1e-05,
                  min_prevalence = 0.2,    
                  transform = "AST",
                  #fixed_effects = c("Age"),
                  random_effects = NULL,
                  normalization = "TSS",
                  analysis_method = "LM",
                  max_significance = 0.25,
                  correction = "BH",
                  standardize = TRUE,
                  cores = 1,
                  plot_heatmap = TRUE,
                  plot_scatter = TRUE, 
                  heatmap_first_n = 50)
}

Maaslin2Fun(dataset=ExprSet, taxa="species")
```


### Visualization of correlations between factors and each baterial species
```{r, fig.height=10, fig.width=6}
correlation_species <- fread("../../Result/MaAsLin2/species/all_results.tsv")

CorPlotFun <- function(dataset=correlation_species,
                       FDR_cut=0.0001,
                       coef_cut=0.002){
  
  # dataset=correlation_species
  # FDR_cut=0.0001
  # coef_cut=0.002
  
  dat <- dataset %>% filter(qval < FDR_cut) %>%
    filter(abs(coef) > coef_cut)
  
  pl <- ggplot(dat, aes(x=metadata, y=feature))+
    geom_tile(aes(fill=coef))+
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
     midpoint = 0, limit = c(-0.05, 0.05), space = "Lab", 
     name="Correlation Coefficient")+
    scale_x_discrete(position="top")+
    labs(x="", y="")+
    guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
          title.position = "bottom", title.hjust = 0.5))+ 
    theme_bw()+
    theme(
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.text.x = element_text(size=10, color="black", angle=90, hjust=0),
      axis.text.y = element_text(size=10, color="black", family="serif", face="italic"),
      text = element_text(size=8, color="black", family="serif"),
      legend.position = "bottom")
  
  return(pl)
}

CorPlotFun(dataset=correlation_species,
           FDR_cut=0.0001,
           coef_cut=0.01)
```


### Systemic information
```{r}
sessionInfo()
```


### Reference 

1. [MaAsLin2](https://huttenhower.sph.harvard.edu/maaslin)

2. [ggplot2 heatmap](http://sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization)
