---
title: "Support Vector Machines on Differential Tax"
date: "Created: 2021-09-18 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introduction

1. Evaluating the important value of protein by recursive feature elimination

2. Adding protein into model when its spearman correlation value with any existing protein in the model was less than 0.7

3. Dividing data into 80% for trainset and 20% for testset

4. Calculating the performance using 10-fold cross-validation

5. To avoid overfitting, Step1-4 was repeated 100 times



Discovery Set: 1111

  * 485 HC
  
  * 145 AA

  * 481 CRC  
  
  
Validation Set: 476 

  * 202 HC
  
  * 62 AA

  * 212 CRC  


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(convert)
library(data.table)
library(caret)
library(kernlab)
library(pROC)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data
```{r}
ExprSet_Discovery <- readRDS("../../Result/Biomarker/RF_model/species_profile_Discovery.RDS")
ExprSet_Validation <- readRDS("../../Result/Biomarker/RF_model/species_profile_Validation.RDS")
CRC_HC_DEA <- fread("../../Result/Biomarker/RF_model/CRC_HC_DEA.csv")
```


### Support Vector Machines
```{r}
SVM_model <- function(dataset=ExprSet_Discovery,
                      testres=CRC_HC_DEA,
                      group_name=subgrp[c(1, 3)],
                      Pname="adj.P.Val",
                      Pval=0.05,
                      LogFC=0,
                      Multicol=FALSE,
                      Rho=0.7,
                      SplitPro=0.8,
                      kfold=5,
                      times=10){
  
  # dataset=ExprSet_Discovery
  # testres=CRC_HC_DEA
  # group_name=subgrp[c(1, 3)]
  # Pname="adj.P.Val"
  # Pval=0.05
  # LogFC=0
  # Multicol=FALSE
  # Rho=0.7
  # SplitPro=0.8
  # kfold=5
  # times=10

  # Differential Tax
  colnames(testres)[which(colnames(testres) == Pname)] <- "PValue"
  
  if(Pval == 0){
    dat_DEP <- testres %>% dplyr::select(GeneID, PValue)
  }else{
    if(LogFC > 0){
      dat_DEP <- testres %>% dplyr::select(GeneID, PValue, logFC) %>%
        filter(PValue < Pval) %>%
        filter(abs(logFC) > LogFC)    
    }else{
      dat_DEP <- testres %>% dplyr::select(GeneID, PValue) %>%
        filter(PValue < Pval)    
    }  
  }
  print(paste0("DE Features: ", dim(dat_DEP)[1]))
  # dataset of Tax
  if(length(group_name) == 3){
    group_name2 <- c("CRC", "non_CRC")
    pheno <- pData(dataset) %>%
      rownames_to_column("SampleID") %>%
      filter(SubGroup%in%group_name) %>%
      mutate(SubGroup=ifelse(SubGroup == group_name2[1], group_name2[1], group_name2[2])) %>%
      mutate(SubGroup=factor(SubGroup, levels = group_name2)) %>%
      column_to_rownames("SampleID")
    group_name <- group_name2
  }else{
    pheno <- pData(dataset) %>%
      rownames_to_column("SampleID") %>%
      filter(SubGroup%in%group_name) %>%
      mutate(SubGroup=factor(SubGroup, levels = group_name)) %>%
      column_to_rownames("SampleID")
  }
  print(table(pheno$SubGroup))
  
  profile <- exprs(dataset)
  edata <- profile[rownames(profile)%in%dat_DEP$GeneID, 
                   colnames(profile)%in%rownames(pheno)] %>% 
    data.frame()
  
  # Data of whole feature 
  mdat <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                       dplyr::select(SampleID, SubGroup),
                     edata %>% t() %>% data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID")
  
  # remove multicollinearity
  if(Multicol){
    feature_cor <- Hmisc::rcorr(as.matrix(mdat[, -1]), type = "spearman")
    index_cor <- findCorrelation(feature_cor$r, cutoff = Rho, verbose = TRUE)
    feature_remain <- feature_cor$r[-index_cor, -index_cor]
    mdat_remain <- mdat %>% dplyr::select(c("SubGroup", rownames(feature_remain)))     
  }else{
    mdat_remain <- mdat
  }
  
  # recursive feature elimination
  require(mlbench)
  rfeControl <- rfeControl(functions=caretFuncs, 
                           method="cv", 
                           number=10, 
                           rerank=TRUE)
  rfebreaks <- seq(2, ncol(mdat_remain), 2)
  rfefit <- rfe(mdat_remain[, -1],
                mdat_remain$SubGroup,
                sizes=rfebreaks,
                rfeControl=rfeControl,
                method="svmRadial")
  
  rfeAccuracy <- tibble(Accuracy = signif(rfefit$results$Accuracy, 3),
                        Variables = rfefit$results$Variables) %>%
                 mutate(label=ifelse(Accuracy == max(Accuracy),
                           paste0("(Features=", Variables, ", Accuracy=", Accuracy,")"), NA))
  rfepl <- ggplot(data=rfeAccuracy, aes(x=Variables, y=Accuracy))+
            geom_point(color="grey", size=3, shape=19)+
            geom_line(color="black", linetype=1, size=1)+
            geom_text(aes(label=label), nudge_y=0.002)+
            annotate(geom="point",
                     x=rfeAccuracy[grep(max(rfeAccuracy$Accuracy), rfeAccuracy$Accuracy), ]$Variables,
                     y=max(rfeAccuracy$Accuracy), color="red", size=4)+
            labs(x="Features (Numbers)",
                 y="Accuracy (Bootstrap)")+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color='black'),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))
  
  # importance value 
  rfeImp <- varImp(rfefit) %>% 
    rownames_to_column("feature") %>%
    filter(feature%in%rfefit$optVariables) %>%
    arrange(desc(Overall))
  mdat_remain_v2 <- mdat_remain %>% dplyr::select(c("SubGroup", rfeImp$feature))
  
  print(paste0("Features after RFE: ", dim(mdat_remain_v2)[2] - 1))
  
  # Data Partition
  #set.seed(123)
  if(!is.null(SplitPro)){
    require(sampling)
    #set.seed(123)
    sample <- sampling::strata(mdat_remain_v2, 
                       stratanames = "SubGroup", 
                       size = rev(round(as.numeric(table(mdat_remain_v2$SubGroup)) * SplitPro)),
                       method = "srswor")
    trainData <- mdat_remain_v2[sample$ID_unit, ]
    testData <- mdat_remain_v2[-sample$ID_unit, ]
  }else{
    trainData <- mdat_remain_v2
    testData <- mdat_remain_v2
  }
  
  # normalization
  mdat_preProcess <- preProcess(mdat_remain_v2, method = c("center", "scale"))
  trainTransformed <- predict(mdat_preProcess, trainData)
  testTransformed <- predict(mdat_preProcess, testData)
  
  # dat_table <- as.matrix(trainTransformed %>% dplyr::select(-SubGroup) %>% data.frame())
  # plot(density(dat_table))
  
  # construct model
  if(dim(trainTransformed)[1] > 10 * 8){
    kfold <- 10
  }else{
    kfold <- kfold
  }
  # N-repeat K-fold cross-validation
  myControl <- trainControl(method = "repeatedcv", 
                           number = kfold,
                           repeats = times,
                           summaryFunction = twoClassSummary,
                           classProbs = TRUE,
                           verboseIter = TRUE)
  
  if(oneFit){
    fit <- train(SubGroup ~., 
                 data = trainTransformed,
                 method = "rf",
                 trControl = myControl,
                 tuneLength = 15,
                 metric = "ROC",
                 verbose = FALSE) 
    # Accuracy
    fit_Acc <- tibble(Feartures = fit$results$mtry,
                      ROC = signif(fit$results$ROC, 3),
                      Sens = signif(fit$results$Sens, 3),
                      Spec = signif(fit$results$Spec, 3)) %>%
                   mutate(label=ifelse(ROC == max(ROC), 
                              paste0("(Features=", Feartures, "\n ROC=", ROC,")"), NA))
    fit_final <- fit
    fit_Acc_final <- fit_Acc
    testData_final <- testTransformed    
    
  }else{
    # adding feature one by one according to their ranked importance value
    fit <- list()
    fit_Acc <- list()
    testData_list <- list()
    j <- 0
    for(i in 3:ncol(trainTransformed)){
      trainTransformed_cln <- trainTransformed[, c(1:i)]
      testTransformed_cln <- testTransformed[, c(1:i)]
      if(all(j == 0)){
        trainTransformed_cln_v2 <- trainTransformed_cln
        testTransformed_cln_v2 <- testTransformed_cln       
      }else{
        trainTransformed_cln_v2 <- trainTransformed_cln[, -j]
        testTransformed_cln_v2 <- testTransformed_cln[, -j]        
      }    
      train_cor <- Hmisc::rcorr(as.matrix(trainTransformed_cln_v2[, -1]), type = "spearman")
      train_index_cor <-  findCorrelation(train_cor$r, cutoff = Rho, verbose = TRUE)
      if(length(train_index_cor) == 0){
        # model 
        fit[[i-2]] <- train(SubGroup ~., 
                      data = trainTransformed_cln_v2,
                      method = "svmRadial",
                      trControl = myControl,
                      tuneLength = 9,
                      metric = "ROC",
                      verbose = FALSE)
        # ROC
        fit_Acc[[i-2]] <- tibble(fit[[i-2]]$results) %>%
                   mutate(ROC=signif(ROC, 3), 
                          label=ifelse(ROC == max(ROC), 
                              paste0("(C_index=", C, "\n ROC=", ROC,")"), NA)) %>%
                   mutate(Model_N=i-2)
        # testData
        testData_list[[i-2]] <- testTransformed_cln_v2
      }else{
        j <- c(j, i)
      }
    }
   
    # summary models
    # if(length(fit) >= 2){
    #   print(summary(resamples(fit)))
    # }
  
    # best model
    fit_Acc_All <- do.call(rbind, fit_Acc) %>%
      arrange(desc(ROC))
    
    fit_final <- fit[[fit_Acc_All$Model_N[1]]]
    fit_Acc_final <- fit_Acc[[fit_Acc_All$Model_N[1]]]
    testData_final <- testData_list[[fit_Acc_All$Model_N[1]]]    
  }  
  
  # the best Accuracy with optimal features
  fit_Acc_pl <- ggplot(data=fit_Acc_final, aes(x=C, y=ROC))+
            geom_point(color="grey", size=3, shape=19)+
            geom_line(color="black", linetype=1, size=1)+
            geom_text(aes(label=label), nudge_y=-0.001)+
            annotate(geom="point",
                     x=fit_Acc_final[grep(max(fit_Acc_final$ROC),
                                              fit_Acc_final$ROC), ]$C,
                     y=max(fit_Acc_final$ROC), color="red", size=4) +
            labs(x="C-index",
                 y="Accuracy (Repeated Cross-Validation)")+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color='black'),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))
  
  # evaluate the performance of model
  pred_raw <- predict(fit_final, newdata = testData_final, type = "raw")
  print(confusionMatrix(pred_raw, testData_final$SubGroup))
  pred_prob <- predict(fit_final, newdata = testData_final, type = "prob")  
  
  # define roc and calculate AUC
  rocobj <- roc(testData_final$SubGroup, pred_prob[, 1])
  auc <- round(auc(testData_final$SubGroup, pred_prob[, 1]), 3)
  roc <- tibble(tpr=rocobj$sensitivities,
                fpr=1 - rocobj$specificities)
  roc_pl <- ggplot(data=roc, aes(x=fpr, y=tpr))+
            geom_path(color="red", size=1)+
            geom_abline(intercept=0, slope=1, color="grey", size=1, linetype=2)+
            labs(x = "False Positive Rate (1 - Specificity)",
                 y = "True Positive Rate (Sensivity or Recall)")+
            annotate("text", x=.75, y=.25, label=paste("AUC =", auc),
                     size=5, family="serif")+
            coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color="black"),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))
  
  # important features
  imp <- fit_final$coefnames
  
  # results
  res <- list(pl=list(rfe=rfepl, acc=fit_Acc_pl, roc=roc_pl),
              model=list(rfe=rfefit, Repeatfit=fit, fit_acc=fit_Acc, fit=fit_final),
              feature=imp)
  return(res)
}

# validation
model_validation <- function(dataset=ExprSet_Validation,
                             datfit=CRC_HC_RF$model$fit,
                             group_name=subgrp[c(1, 3)]){
  
  # dataset=ExprSet_Validation
  # datfit=CRC_HC_RF$model$fit
  # group_name=subgrp[c(1, 3)]
  
  # dataset of DEP
  if(length(group_name) == 3){
    group_name2 <- c("CRC", "non_CRC")
    pheno <- pData(dataset) %>%
      rownames_to_column("SampleID") %>%
      filter(SubGroup%in%group_name) %>%
      mutate(SubGroup=ifelse(SubGroup == group_name2[1], group_name2[1], group_name2[2])) %>%
      mutate(SubGroup=factor(SubGroup, levels = group_name2)) %>%
      column_to_rownames("SampleID")
    group_name <- group_name2
  }else{
    pheno <- pData(dataset) %>%
      rownames_to_column("SampleID") %>%
      filter(SubGroup%in%group_name) %>%
      mutate(SubGroup=factor(SubGroup, levels = group_name)) %>%
      column_to_rownames("SampleID")
  }
  
  profile <- exprs(dataset)
  edata <- profile[, colnames(profile)%in%rownames(pheno)] %>% 
    data.frame()
  
  # Data of whole feature 
  mdat <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                       dplyr::select(SampleID, SubGroup),
                     edata %>% t() %>% data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID")
  
  # evaluate the performance of model
  pred_raw <- predict(datfit, newdata = mdat, type = "raw")
  print(confusionMatrix(pred_raw, mdat$SubGroup))
  pred_prob <- predict(datfit, newdata = mdat, type = "prob")  
  
  # define roc and calculate AUC
  rocobj <- roc(mdat$SubGroup, pred_prob[, 1])
  auc <- round(auc(mdat$SubGroup, pred_prob[, 1]), 3)
  roc <- tibble(tpr=rocobj$sensitivities,
                fpr=1 - rocobj$specificities)
  roc_pl <- ggplot(data=roc, aes(x=fpr, y=tpr))+
            geom_path(color="red", size=1)+
            geom_abline(intercept=0, slope=1, color="grey", size=1, linetype=2)+
            labs(x = "False Positive Rate (1 - Specificity)",
                 y = "True Positive Rate (Sensivity or Recall)")+
            annotate("text", x=.75, y=.25, label=paste("AUC =", auc),
                     size=5, family="serif")+
            coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color="black"),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))
  
  # Accuracy
  CFM <- confusionMatrix(pred_raw, mdat$SubGroup)
  CFM_index <- data.frame(Value=CFM$byClass) %>%
    rownames_to_column("Index") %>% 
    rbind(data.frame(Index="Accuracy", Value=as.numeric(CFM$overall[1])))
  
  res <- list(auc=auc, roc=roc_pl, index=CFM_index)
  
  return(res)
}

if(!dir.exists("../../Result/Biomarker/SVM_model/")){
  dir.create("../../Result/Biomarker/SVM_model/", recursive = T)
}
```


### 10-repeat 10-fold cross validation random forest
```{r}
# if(!file.exists("../../Result/Biomarker/SVM_model/CRC_HC_SVM_Discovery.RDS")){
#    CRC_HC_SVM <- SVM_model(
#                      dataset=ExprSet_Discovery,
#                      testres=CRC_HC_DEA,
#                      group_name=subgrp[c(1, 3)],
#                      Pname="adj.P.Val",
#                      Pval=0.05,
#                      LogFC=0,
#                      Multicol=FALSE,
#                      Rho=0.7,
#                      SplitPro=0.7,
#                      kfold=5,
#                      times=10,
#                      oneFit=TRUE)
#   saveRDS(CRC_HC_SVM, "../../Result/Biomarker/SVM_model/CRC_HC_SVM_Discovery.RDS", compress = TRUE)
# }else{
#   CRC_HC_SVM <- readRDS("../../Result/Biomarker/SVM_model/CRC_HC_SVM_Discovery.RDS")
# }


CRC_HC_SVM <- SVM_model(
                     dataset=ExprSet_Discovery,
                     testres=CRC_HC_DEA,
                     group_name=subgrp[c(1, 3)],
                     Pname="adj.P.Val",
                     Pval=0.05,
                     LogFC=0,
                     Multicol=FALSE,
                     Rho=0.7,
                     SplitPro=0.7,
                     kfold=5,
                     times=10,
                     oneFit=TRUE)
CRC_HC_SVM_validation <- model_validation(
                       dataset=ExprSet_Validation,
                       datfit=CRC_HC_SVM$model$fit,
                       group_name=subgrp[c(1, 3)])

# while(with(CRC_HC_SVM_validation, auc < 0.8 | index$Value[1] < 0.7 | index$Value[2] < 0.7))
while(with(CRC_HC_SVM_validation, index$Value[12] < 0.7)){
  CRC_HC_SVM <- SVM_model(
                     dataset=ExprSet_Discovery,
                     testres=CRC_HC_DEA,
                     group_name=subgrp[c(1, 3)],
                     Pname="adj.P.Val",
                     Pval=0.05,
                     LogFC=0,
                     Multicol=FALSE,
                     Rho=0.7,
                     SplitPro=0.7,
                     kfold=5,
                     times=10,
                     oneFit=TRUE)
  CRC_HC_SVM_validation <- model_validation(
                 dataset=ExprSet_Validation,
                 datfit=CRC_HC_SVM$model$fit,
                 group_name=subgrp[c(1, 3)])
}

saveRDS(CRC_HC_SVM, "../../Result/Biomarker/SVM_model/CRC_HC_SVM_Discovery.RDS", compress = TRUE)
```



### systemic information
```{r}
sessionInfo()
```


### Reference

1. [Underdevelopment of the gut microbiota and bacteria species as non-invasive markers of prediction in children with autism spectrum disorder](https://pubmed.ncbi.nlm.nih.gov/34312160/)

2. [Recursive Feature Elimination](https://topepo.github.io/caret/recursive-feature-elimination.html)

3. [The 5th Tribe, Support Vector Machines and caret](https://www.r-bloggers.com/2015/10/the-5th-tribe-support-vector-machines-and-caret/)

4. [SUPPORT VECTOR MACHINE CLASSIFIER IMPLEMENTATION IN R WITH CARET PACKAGE](https://dataaspirant.com/support-vector-machine-classifier-implementation-r-caret-package/)

