---
title: "Functional module was contributed by a mixture of species: stacked barplot"
date: "Created: 2021-10-18 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introduction 

The significant taxa were identified by hypothesis test in *pathway.Rmd* (`CRC_nonCRC_DEA %>% filter(Enrichment=="CRC") %>% select(FeatureID)`)

1. P162-PWY:_L-glutamate_degradation_V_(via_hydroxyglutarate)
2. PWY-6263:_superpathway_of_menaquinol-8_biosynthesis_II
3. PWY-7288:_fatty_acid_&beta;-oxidation_(peroxisome,_yeast)
4. PWY-6318:_L-phenylalanine_degradation_IV_(mammalian,_via_side_chain)
5. PWY-6285:_superpathway_of_fatty_acids_biosynthesis_(E._coli)
6. PWY-5850:_superpathway_of_menaquinol-6_biosynthesis_I
7. PWY-5896:_superpathway_of_menaquinol-10_biosynthesis
8. PWY-5860:_superpathway_of_demethylmenaquinol-6_biosynthesis_I
9. PWY-7204:_pyridoxal_5'-phosphate_salvage_II_(plants)
10. PWY-5920:_superpathway_of_heme_biosynthesis_from_glycine
11. PWY-5747:_2-methylcitrate_cycle_II
12. PWY-5675:_nitrate_reduction_V_(assimilatory)
13. PWY-4702:_phytate_degradation_I
14. PWY-6749:_CMP-legionaminate_biosynthesis_I
15. PWY-561:_superpathway_of_glyoxylate_cycle_and_fatty_acid_degradation
16. KETOGLUCONMET-PWY:_ketogluconate_metabolism


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(convert)
library(ggplot2)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
top.col <- c("#010281", "#0428FF", "#00D2F2",
             "#77FA72", "#FFEA08", "#FE4904",
             "#810402", "#7E7E7E", "#CECECE")
```


### Input data
```{r}
ExprSet <- readRDS("../../Result/Profile/pathway_taxa_ExprSet.RDS")
```


### Function
```{r}
BarPotFun <- function(datset=ExprSet,
                      pathwaylist="PWY-7204:_pyridoxal_5'-phosphate_salvage_II_(plants)",
                      group_info="SubGroup",
                      group_name=subgrp[c(3,2,1)],
                      group_col=grp.col[c(3,1)],
                      ntop=7,
                      abundance_cutoff=1e-07){
  
  # datset=ExprSet
  # pathwaylist="PWY-7204:_pyridoxal_5'-phosphate_salvage_II_(plants)"
  # group_info="SubGroup"
  # group_name=subgrp[c(3,2,1)]
  # group_col=grp.col[c(3,1)]
  # ntop=7
  # abundance_cutoff=1e-07
  
  
  phenotype <- pData(datset)
  profile <- exprs(datset)
  feature <- fData(datset)
  
  feature_cln <- feature %>% filter(pathwayID%in%pathwaylist)
  profile_cln <- profile[rownames(profile)%in%feature_cln$featureID, ] %>% data.frame()
  feature_cln[is.na(feature_cln)] <- "Unclassified"
  rownames(profile_cln) <- feature_cln$species
  
  colnames(phenotype)[which(colnames(phenotype) == group_info)] <- "Group"
  if(length(group_name) == 3){
    group_name2 <- c("CRC", "non_CRC")
    pheno <- phenotype %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[1], group_name2[1], group_name2[2])) %>%
      mutate(Group=factor(Group, levels = group_name2))
    group_name <- group_name2
  }else{
    pheno <- phenotype %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name))
  }  
  edata <- profile_cln[, colnames(profile_cln)%in%pheno$SampleID]
  
  mdat <- inner_join(pheno %>% dplyr::select(SampleID, Group),
                     edata %>% t() %>% 
                       data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") 
  
  # converting species into other and unclassified kind
  mdat_mean <-  mdat %>%
    dplyr::select(-c("SampleID", "Group")) %>%
    summarise(across(everything(), mean)) %>%
    tidyr::gather(key="taxa", value="value") %>%
    arrange(desc(value))
  
  mdat_mean_unclassified <- mdat_mean %>% filter(taxa=="Unclassified") %>%
    mutate(Taxa_Name="Unclassified")
  mdat_mean_top <- mdat_mean %>% filter(taxa!="Unclassified") %>%
    dplyr::slice(c(1:ntop)) %>%
    mutate(Taxa_Name=taxa)
  mdat_mean_Other <- mdat_mean %>% filter(taxa!="Unclassified") %>%
    dplyr::slice(-c(1:ntop)) %>%
    mutate(Taxa_Name="Other")
  mdat_mean_final <- rbind(mdat_mean_top, mdat_mean_Other, mdat_mean_unclassified)
  
  # plotdata
  plotdata_prepare <- mdat %>%
      dplyr::select(-Group) %>%
      tidyr::gather(key="taxa", value="value", -SampleID) %>%
      inner_join(mdat_mean_final %>% dplyr::select(taxa, Taxa_Name), by="taxa") %>%
      inner_join(mdat %>% dplyr::select(SampleID, Group), by = "SampleID") %>%
      mutate(Taxa_group=factor(Taxa_Name, levels = unique(mdat_mean_final$Taxa_Name)))
  
  temp_rank <- plotdata_prepare %>% dplyr::select(SampleID, Group, value) %>%
    group_by(Group, SampleID) %>%
    summarise(Sum=sum(value)) %>%
    arrange(Group, desc(Sum)) %>%
    filter(Sum > abundance_cutoff)
  
  plotdata <- plotdata_prepare %>% 
    filter(SampleID%in%temp_rank$SampleID) %>%
    mutate(SampleID=factor(SampleID, levels = c(temp_rank$SampleID)))
  
  pl <- ggplot(plotdata, aes(x=SampleID, y=value, fill=Taxa_group))+
          geom_bar(stat="identity", position="stack")+
          scale_y_continuous(expand=c(0, 0))+
          scale_fill_manual(values=top.col, name="Stratifications")+
          labs(x=paste0("Samples(N=", nrow(temp_rank), ")"), y="Relative abundance",
               title=pathwaylist)+
          facet_wrap(facets = "Group", ncol=2, scales="free_x")+
          theme_bw()+
          theme(plot.title = element_text(face="bold", color="black", size = 14, hjust = .5),
                axis.title = element_text(face="bold", color="black", size = 12),
                axis.text.y = element_text(color="black", size=10),
                axis.text.x = element_blank(),
                axis.ticks.x = element_blank(),
                text = element_text(size=8, color="black", family="serif"),
                panel.grid = element_blank(),
                legend.position = "right",
                legend.key.height = unit(0.6,"cm"),
                legend.text = element_text(face="italic", color="black", size=10),
                strip.text = element_text(size=10, color="black", face="bold")) 
  
  return(pl)
}
```


### CRC vs nonCRC
```{r, fig.width=10, fig.height=5}
CRC_nonCRC_BarPot <- BarPotFun(
                        datset=ExprSet,
                        pathwaylist="PWY-7204:_pyridoxal_5'-phosphate_salvage_II_(plants)",
                        group_info="SubGroup",
                        group_name=subgrp[c(3,2,1)],
                        group_col=grp.col[c(3,1)])
CRC_nonCRC_BarPot
```


### systemic information
```{r}
sessionInfo()
```
