---
title: "Visualization of Differential Expression Species"
date: "Created: 2021-11-10 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introdution

DataStatus: 

  * Healthy Control: HC
  
  * Adenoma: AA	
  
  * Colorectal Cancer: CRC
  


* Discovery Cohort: 1112

  * 626 non-CRC samples (481 HC + 145 AA)
  
  * 486 CRC samples

* Validation Cohort: 475 

  * 268 non-CRC samples(206 HC + 62 AA)
  
  * 207 CRC samples
  
  

Three comparisons between the pancreatic tumor (PDAC) patients and other groups:

  1. CRC 486 vs HC 481
  
  2. CRC 486 vs AA 145
  
  3. AA 145 vs HC 481
  
  4. CRC 486 vs non-CRC( 481 HC; 145 AA) 626


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(convert)
library(data.table)
library(ggplot2)
library(ggrepel)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data 
```{r}
ExprSet <- readRDS("../../Result/Biomarker/Logistic_model/species_profile_Discovery.RDS")
CRC_nonCRC_DEA <- fread("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_final.csv")
```


### Function
```{r}
VolcanoFun <- function(datset=CRC_nonCRC_DEA,
                       genelist=c("Asaccharobacter_celatus", "Odoribacter_splanchnicus"),
                       group_name=subgrp[c(3,2,1)],
                       group_col=grp.col[c(3,1)],
                       Pval=0.05,
                       LogFC=0.5){

  # datset=CRC_nonCRC_DEA
  # genelist=c("Asaccharobacter_celatus", "Odoribacter_splanchnicus")
  # group_name=subgrp[c(3,2,1)]
  # group_col=grp.col[c(3,1)]
  # Pval=0.05
  # LogFC=0.5
  
  if(length(group_name) == 3){
      group_name2 <- c("CRC", "non_CRC")
      group_name <- group_name2
    }  

  dat <- datset %>%
    mutate(color=factor(Enrichment,
                        levels = c(group_name, "Nonsignif"))) %>%
    mutate(color_col=ifelse(color == group_name[1], group_col[1],
                            ifelse(color == group_name[2], group_col[2], "grey80")))
  # print(table(dat$color))
  dat_status <- table(dat$color)
  dat_status_number <- as.numeric(dat_status)
  dat_status_name <- names(dat_status)
  legend_label <- c(paste0(dat_status_name[1], " (", dat_status_number[1], ")"),
                    paste0(dat_status_name[2], " (", dat_status_number[2], ")"),
                    paste0("Nonsignif", " (", dat_status_number[3], ")"))

  dat.signif <- subset(dat, adj.P.Val < Pval & abs(logFC) > LogFC) %>%
    filter(FeatureID%in%genelist)
  print(table(dat.signif$color))

  group_col_new <- c(rev(group_col), "grey80")
  group_name_new <- levels(dat$color)

  xlabel <- paste0("log2(", paste(group_name, collapse="/"), ")")

  # Make a basic ggplot2 object with x-y values
  pl <- ggplot(dat, aes(x=logFC, y=-log10(adj.P.Val), color=color))+
          geom_point(size=2, alpha=1, stroke=1)+
          # geom_point(aes(fill=color), size=2, shape=21, stroke=.5, color="grey")+
          # scale_fill_manual(name = NULL,
          #                    values = group_col_new,
          #                    labels = c(legend_label, "Nonsignif"))+
          scale_color_manual(name=NULL,
                             values=group_col_new,
                             labels=c(legend_label, "Nonsignif"))+    
          xlab(xlabel) +
          ylab(expression(-log[10]("adjusted p-value")))+
          geom_hline(yintercept=-log10(Pval), alpha=.8, linetype=2, size=.7)+
          geom_vline(xintercept=LogFC, alpha=.8, linetype=2, size=.7)+
          geom_vline(xintercept=-LogFC, alpha=.8, linetype=2, size=.7)+
          # geom_text_repel(data = dat.signif,
          #                 aes(fill = color_col, label = FeatureID, segment.color = color_col),
          #                 size = 4,
          #                 point.padding = 0,
          #                 min.segment.length = 0,
          #                 max.time = 1,
          #                 max.iter = 1e5,
          #                 box.padding = 0.5,
          #                 color = "white",
          #                 bg.color = "grey30",
          #                 bg.r = 0.15)+    
    
          geom_text_repel(data = dat.signif,
                          aes(fill = color, label = FeatureID),
                          size = 5,
                          max.overlaps = getOption("ggrepel.max.overlaps", default = 80),
                          segment.linetype = 1,
                          segment.curvature = -1e-20,
                          segment.color = "grey50",
                          box.padding = unit(0.35, "lines"),
                          point.padding = unit(0.3, "lines"),
                          arrow = arrow(length = unit(0.005, "npc")),
                          face = "bold",
                          # color = "black",     # text color
                          # bg.color = "white",  # shadow color
                          bg.r = 0.15)+
          annotate("text", x=min(dat$logFC), y=-log10(Pval), label=Pval, size=6, color="red")+
          annotate("text", x=LogFC, y=0, label=LogFC, size=6, color="red")+
          annotate("text", x=-LogFC, y=0, label=-LogFC, size=6, color="red")+
          scale_y_continuous(trans = "log1p")+
          guides(color=guide_legend(override.aes = list(size = 3)))+
          theme_bw()+
          theme(axis.title = element_text(color = "black", size = 12),
                axis.text = element_text(color = "black", size = 10),
                text = element_text(size = 8, color = "black", family="serif"),
                panel.grid = element_blank(),
                #legend.position = "right",
                legend.position = c(.15, .1),
                legend.key.height = unit(0.6,"cm"),
                legend.text = element_text(face = "bold", color = "black", size = 8),
                strip.text = element_text(face = "bold", size = 14))
  return(pl)
}

BoxplotFun <- function(datset=ExprSet,
                       genelist=c("Asaccharobacter_celatus", "Odoribacter_splanchnicus"),
                       group_name=subgrp[c(3,2,1)],
                       group_col=grp.col[c(3,1)]){

  # datset=ExprSet
  #     genelist=c("Asaccharobacter_celatus", "Odoribacter_splanchnicus")
  # group_name=subgrp[c(3,2,1)]
  # group_col=grp.col[c(3,1)]

  if(length(group_name) == 3){
    group_name2 <- c("CRC", "non_CRC")
    pheno <- pData(datset) %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[1], group_name2[1], group_name2[2])) %>%
      mutate(Group=factor(Group, levels = group_name2)) %>%
      column_to_rownames("SampleID")
    group_name <- group_name2
  }else{
    pheno <- pData(datset) %>%
      rownames_to_column("SampleID") %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name)) %>%
      column_to_rownames("SampleID")
  }
  print(table(pheno$Group))  

  edata <- data.frame(exprs(datset)) %>%
            dplyr::select(rownames(pheno)) %>%
            rownames_to_column("FeatureID") %>%
            filter(FeatureID%in%genelist) %>%
            column_to_rownames("FeatureID")

  mdat <- pheno %>% dplyr::select(Group) %>%
    rownames_to_column("SampleID") %>%
    inner_join(t(edata) %>% data.frame() %>% rownames_to_column("SampleID"), by = "SampleID") %>%
    column_to_rownames("SampleID")

  plotdata <- mdat %>% tidyr::gather(key="FeatureID", value="value", -Group) %>%
    mutate(Group=factor(Group, levels = group_name))
  plotdata$FeatureID <- factor(plotdata$FeatureID, levels = genelist)

  pl <- ggplot(plotdata, aes(x=Group, y=value, fill=Group))+
          stat_boxplot(geom="errorbar", width=0.15,
                       position=position_dodge(0.4)) +
          geom_boxplot(width=0.4,
                       outlier.colour="black",
                       outlier.shape=21,
                       outlier.size=.5)+
          scale_fill_manual(values=group_col)+
          facet_wrap(facets="FeatureID", scales="free_y")+
          labs(x="", y="Relative Abundance")+
          guides(fill="none")+
          theme_classic()+
          theme(axis.title = element_text(color="black", size=12),
                axis.text.x = element_text(color="black", size=10, hjust=.5, vjust=.5, angle=60),
                text = element_text(size=8, color="black", family="serif"),
                panel.grid = element_blank(),
                strip.text = element_text(face="bold", size=12))

  return(pl)
}
```


### Biomarkers
```{r, fig.width=8, fig.height=6}
CRC_nonCRC_Volcano <- VolcanoFun(datset=CRC_nonCRC_DEA,
                                 group_name=subgrp[c(3,2,1)],
                                 group_col=grp.col[c(3,1)])
CRC_nonCRC_Volcano

CRC_nonCRC_Boxplot <- BoxplotFun(datset=ExprSet,
                                 group_name=subgrp[c(3,2,1)],
                                 group_col=grp.col[c(3,1)])
CRC_nonCRC_Boxplot
```


### systemic information
```{r}
sessionInfo()
```
