---
title: "LASSO Logistic regression model to calculate Proportion Score"
date: "Created: 2021-11-10 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introdution

Building LASSO regression model based on the Differential Expression Proteins and Calculating the Proportion Score. The procedures:

1. Randomly split discovery cohort into trainset and testset with 0.7 probability

2. The logistic LASSO regression model was fitted to the trainset

3. The fitted model was employed on testset 

This process was repeated 100 times, thus generating 100 different logistic LASSO regression models.

========================================================================

DataStatus: 

  * Healthy Control: HC
  
  * Adenoma: AA	
  
  * Colorectal Cancer: CRC

* Discovery Cohort: 1112

  * 626 non-CRC samples (481 HC + 145 AA)
  
  * 486 CRC samples

========================================================================

General Evaluation of model performance

* AUC

* Accuracy

* Sensitivity

* Specificity 

* Positive Predictive Value (Precision)

========================================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
library(dplyr)
library(tibble)
library(convert)
library(data.table)
library(glmnet)
library(ggplot2)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data 
```{r}
ExprSet <- readRDS("../../Result/Biomarker/Logistic_model/species_profile_Discovery.RDS")

CRC_HC_DEA <- fread("../../Result/Biomarker/Logistic_model/CRC_HC_DEA_final.csv")
CRC_nonCRC_DEA <- fread("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_final.csv")
```


### logistic LASSO regression model & Proportion score
```{r}
build_LASSO <- function(dataset=ExprSet,
                        testres=CRC_nonCRC_DEA,
                        group_name=subgrp[1:3],
                        Pname="adj.P.Val",
                        Pval=0.05,
                        LogFC=0,
                        splitPro=0.7,
                        kfold=10,
                        times=100){
    
  # dataset=ExprSet
  # testres=CRC_nonCRC_DEA
  # group_name=subgrp[1:3]
  # Pname="adj.P.Val"
  # Pval=0.05
  # LogFC=0
  # splitPro=0.7
  # kfold=10
  # times=10
  
  
  # Differential features
  colnames(testres)[which(colnames(testres) == Pname)] <- "PValue"
  if(Pval == 0){
    dat_DEP <- testres %>% dplyr::select(FeatureID, PValue)
  }else{
    if(LogFC > 0){
      dat_DEP <- testres %>% dplyr::select(FeatureID, PValue, logFC) %>%
        filter(PValue < Pval) %>%
        filter(abs(logFC) > LogFC)    
    }else{
      dat_DEP <- testres %>% dplyr::select(FeatureID, PValue) %>%
        filter(PValue < Pval)    
    }  
  }
  print(dim(dat_DEP)[1])
  
  # dataset of DEP
  if(length(group_name) == 3){
    group_name2 <- c("non_CRC", "CRC")
    pheno <- pData(dataset) %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[2], group_name2[2], group_name2[1])) %>%
      mutate(Group=factor(Group, levels = group_name2)) 
    group_name <- group_name2
  }else{
    pheno <- pData(dataset) %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name))    
  }
  edata <- exprs(dataset)[dat_DEP$FeatureID, rownames(pheno)] %>% data.frame()
  
  mdat_temp <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                       dplyr::select(SampleID, Group),
                     edata %>% t() %>% data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID") 
  
  # Before building model, performing Zscore standardization
  mdat <- cbind(mdat_temp$Group, scale(mdat_temp %>% dplyr::select(-Group)) %>% data.frame())
  colnames(mdat)[1] <- "Group"  
  
  # scale 
  LASSO_model <- list()
  auc_score <- c()
  # repeat building model
  for(i in 1:times){
    # Data Partition
    #set.seed(123)
    require(sampling)
    sample <- sampling::strata(mdat, 
                     stratanames = "Group",
                     size = round(as.numeric(table(mdat$Group)) * splitPro),
                     method = "srswor")
    trainData <- mdat[sample$ID_unit, ]
    testData <- mdat[-sample$ID_unit, ]
    
    dat_table <- as.matrix(trainData %>% dplyr::select(-Group))
    dat_target <- trainData$Group
    print("The Group's number of TrainData")
    print(table(trainData$Group))    
    
    # preview data distribution
    #plot(density(dat_table))
    
    # determine folds
    if(dim(trainData)[1] > 10 * 8){
      kfold <- 10
    }else{
      kfold <- kfold
    }     
    
    # feature selection
    #set.seed(123)  
    cvfit <- cv.glmnet(
                     x=dat_table,
                     y=dat_target,
                     family="binomial",
                     nfolds = kfold,
                     alpha = 1,
                     nlambda = 100,
                     type.measure = "auc")
    #plot(cvfit)
    predicted <- predict(cvfit, as.matrix(testData[, -1]), type = "response", s = cvfit$lambda.min)
    
    predict_class <- predict(cvfit, as.matrix(testData[, -1]), type = "class")
    print(caret::confusionMatrix(factor(predict_class), factor(testData$Group)))
    
    # AUC
    require(ggplot2)
    require(pROC)
    rocobj <- roc(testData$Group, predictor=predicted[, 1])
    
    # AUC 95% CI 
    auc_95CI <- ci.auc(rocobj, conf.level = 0.95)
    auc_95CI_value <- round(as.numeric(auc_95CI), 3)
    auc_95CI_label <- paste0("AUC = ", auc_95CI_value[2], " (", 
                              auc_95CI_value[1], ", ", 
                              auc_95CI_value[3], ")")
  
    auc <- round(auc(testData$Group, predictor=predicted[, 1]), 4)
    pl <- ggroc(rocobj, color="red", linetype=1, size=1, alpha=1, legacy.axes=T)+
                  geom_abline(intercept=0, slope=1, color="grey", size=1, linetype=1)+
                labs(x="False Positive Rate (1 - Specificity)",
                     y="True Positive Rate (Sensivity or Recall)")+
                # annotate("text", x=.75, y=.25, label=paste("AUC =", auc),
                #          size=5, family="serif")+
                annotate("text", x=.75, y=.25, label=auc_95CI_label,
                         size=5, family="serif")+      
                coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
                theme_bw()+
                theme(panel.background = element_rect(fill="transparent"),
                      axis.ticks.length = unit(0.4, "lines"), 
                      axis.ticks = element_line(color="black"),
                      axis.line = element_line(size=.5, color="black"),
                      axis.title = element_text(color="black", size=12, face="bold"),
                      axis.text = element_text(color="black", size=10),
                      text = element_text(size=8, color="black", family="serif"))  
  
    # predictors
    predictors <- data.frame(as.matrix(coef(cvfit, cvfit$lambda.min))) %>%
        setNames("Score") %>%
        rownames_to_column("Protein") %>%
        slice(-1) %>%
        filter(Score!=0) 
    
    auc_score <- c(auc_score, auc) 
    # model results
    LASSO_model[[i]] <- list(fit=cvfit, 
                             auc=auc,
                             auc_CI=auc_95CI_label,
                             pl=pl,
                             marker=predictors)    
  } 
  
  print(summary(auc_score))
  
  return(LASSO_model)
}

# AUC 95% confidence Interval
AUC_95CI_plot <- function(dataset=PDAC_nonPDAC_LASSO){
  
  # dataset=PDAC_nonPDAC_LASSO
  
  # auc_score
  auc_score <- c()
  for (j in 1:length(dataset)) {
    
    temp <- dataset[[j]]$auc
    auc_score <- c(auc_score, temp)
  }
  
  # 95% CI 
  CI95_Normal <- function(x){
    # Calculating a Confidence Interval From a Normal Distribution
    x_length <- length(x)
    x_mean <- signif(mean(x), 3)
    x_sd <- sd(x)
    error <- qnorm(0.975) * x_sd/sqrt(x_length)
    left <- signif(x_mean - error, 3)
    right <- signif(x_mean + error, 3)
    res <- paste0("AUC = ", x_mean, " (", left, ", ", right, ")")
    return(res)
  }
  
  CI95_t <- function(x){
    # Calculating a Confidence Interval From a t Distribution
    x_length <- length(x)
    x_mean <- signif(mean(x), 3)
    x_sd <- sd(x)
    error <- qt(0.975, df=x_length-1) * x_sd/sqrt(x_length)
    left <- signif(x_mean - error, 3)
    right <- signif(x_mean + error, 3)
    res <- paste0("AUC = ", x_mean, " (", left, ", ", right, ")")
    return(res)
  }
  
  auc_95CI_Normal <- CI95_Normal(auc_score)
  
  # plot 
  dat_range <- round(as.numeric(summary(auc_score)), 3)
  dat_name <- names(summary(auc_score))
  pl <- ggplot(data.frame(x=auc_score), aes(x=x)) + 
          geom_density()+
          labs(x="AUC", y="density")+
          annotate(geom="text", x=(dat_range[1]+dat_range[2])/2, y=5, label=auc_95CI_Normal, size=6)+
          geom_vline(xintercept=dat_range[2], linetype=2)+
          geom_vline(xintercept=dat_range[5], linetype=2)+
          scale_x_continuous(breaks=dat_range,
                             labels=paste(dat_name, dat_range, sep = "\n"))+
          theme_bw()+
          theme(panel.background = element_rect(fill="transparent"),
                axis.ticks.length = unit(0.4, "lines"), 
                axis.ticks = element_line(color="black"),
                axis.line = element_line(size=.5, color="black"),
                axis.title = element_text(color="black", size=12, face="bold"),
                axis.text.x = element_text(angle = 90, hjust=1, face="bold"),
                axis.text = element_text(color="black", size=10),
                text = element_text(size=8, color="black", family="serif")) 
  
  return(pl)
}

# Protein signature
ProportionScore <- function(datfit=PDAC_nonPDAC_LASSO){
  
  # datfit=PDAC_nonPDAC_LASSO
  
  df_marker <- data.frame()
  for (i in 1:length(datfit)){
    temp <- datfit[[i]]$marker
    df_marker <- rbind(df_marker, temp)
  }
  
  # proportion score
  df_marker_proportion <- df_marker %>% 
    group_by(Protein) %>%
    summarise(Count=length(Protein)) %>%
    mutate(Proportion=Count/length(datfit))
  
  # protein signatures
  range_score_list <- list()
  range_score <- seq(0, 1, 0.05)
  for(i in 1:(length(range_score)-1)){
    range_score_list[[i]] <- c(range_score[i], range_score[i+1])
  }
  range_score_list[[length(range_score)]] <- range_score[1]
  range_score_list[[length(range_score)+1]] <- range_score[length(range_score)]
  
  df_marker_proportion$Signature <- NA
  for(j in 1:nrow(df_marker_proportion)){
    for(m in 1:length(range_score_list)){
      if(length(range_score_list[[m]]) > 1){
        df_marker_proportion$Signature[j] <- ifelse(
          df_marker_proportion$Proportion[j] > range_score_list[[m]][1] &
                df_marker_proportion$Proportion[j] <= range_score_list[[m]][2],
                      paste0("Signature", m), df_marker_proportion$Signature[j])
      }else{
        df_marker_proportion$Signature[j] <- ifelse(
          df_marker_proportion$Proportion[j] == range_score_list[[m]][1],
                paste0("Signature", m), df_marker_proportion$Signature[j])
      }
    }
  }
  
  return(df_marker_proportion)
}


if(!dir.exists("../../Result/Biomarker/Logistic_model")){
  dir.create("../../Result/Biomarker/Logistic_model", recursive = T)
}
```


### CRC vs HC
```{r}
CRC_HC_LASSO <- build_LASSO(dataset=ExprSet,
                            testres=CRC_HC_DEA,
                            group_name=subgrp[c(1,3)],
                            Pname="adj.P.Val",
                            Pval=0.05,
                            LogFC=0, 
                            splitPro=0.5,
                            kfold=10,
                            times=100)

CRC_HC_LASSO_plot <- AUC_95CI_plot(dataset=CRC_HC_LASSO)
CRC_HC_LASSO_plot

CRC_HC_DEA_signature <- ProportionScore(datfit=CRC_HC_LASSO)
DT::datatable(CRC_HC_DEA_signature)


CRC_HC_filename_RDS <- paste0("../../Result/Biomarker/Logistic_model/CRC_HC_DEA_LASSO-", Sys.Date(), ".RDS")
saveRDS(CRC_HC_LASSO, file = CRC_HC_filename_RDS, compress = TRUE)

CRC_HC_filename_pscore <- paste0("../../Result/Biomarker/Logistic_model/CRC_HC_DEA_signature-", Sys.Date(), ".csv")
write.csv(CRC_HC_DEA_signature, file = CRC_HC_filename_pscore, row.names = F)
```


### CRC vs non_CRC
```{r}
CRC_nonCRC_LASSO <- build_LASSO(
                              dataset=ExprSet,
                              testres=CRC_nonCRC_DEA,
                              group_name=subgrp[1:3],
                              Pname="adj.P.Val",
                              Pval=0.05,
                              LogFC=0, 
                              splitPro=0.7,
                              kfold=10,
                              times=100)

CRC_nonCRC_LASSO_plot <- AUC_95CI_plot(dataset=CRC_nonCRC_LASSO)
CRC_nonCRC_LASSO_plot

CRC_nonCRC_DEA_signature <- ProportionScore(datfit=CRC_nonCRC_LASSO)
DT::datatable(CRC_nonCRC_DEA_signature)


CRC_nonCRC_filename_RDS <- paste0("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_LASSO-", Sys.Date(), ".RDS")
saveRDS(CRC_nonCRC_LASSO, file = CRC_nonCRC_filename_RDS, compress = TRUE)

CRC_nonCRC_filename_pscore <- paste0("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_signature-", Sys.Date(), ".csv")
write.csv(CRC_nonCRC_DEA_signature, file = CRC_nonCRC_filename_pscore, row.names = F)
```


### systemic information
```{r}
sessionInfo()
```


### Reference

1. [An Introduction to glmnet](https://glmnet.stanford.edu/articles/glmnet.html)

2. [how to calculate AUC confidence interval](https://www.r-bloggers.com/2019/08/how-to-get-an-auc-confidence-interval/)

