---
title: "Logistic Ridge regression model based on the signatures"
date: "Created: 2021-11-10 Updated: `r Sys.Date()`"
author: 
  - name: "Hua Zou"
    email: "zouhua1@outlook.com"
output: 
  html_notebook:
    codes: hide
---

### Introdution


=====================================================

DataStatus: 

  * Healthy Control: HC
  
  * Adenoma: AA	
  
  * Colorectal Cancer: CRC

* Discovery Cohort: 1112

  * 626 non-CRC samples (481 HC + 145 AA)
  
  * 486 CRC samples

* Validation Cohort: 475 

  * 268 non-CRC samples(206 HC + 62 AA)
  
  * 207 CRC samples
  
=======================================================

Four comparisons between the Colorectal Cancer (CRC) patients and other groups:

  1. CRC 486 vs HC 481
  
  2. CRC 486 vs AA 145
  
  3. AA 145 vs HC 481
  
  4. **CRC 486 vs non-CRC( 481 HC; 145 AA) 626**
  
========================================================================

General Evaluation of model performance

* AUC

* Accuracy

* Sensitivity

* Specificity 

* Positive Predictive Value(Precision)

========================================================================


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = T)
library(dplyr)
library(tibble)
library(convert)
library(data.table)
library(glmnet)
library(ggplot2)

# rm(list = ls())
options(stringsAsFactors = F)
options(future.globals.maxSize = 1000 * 1024^2)

grp <- c("control", "adenoma", "CRC")
subgrp <- c("HC", "AA", "CRC")
grp.col <- c("#568875", "#73FAFC", "#EE853D")
```


### Importing Data 
```{r}
ExprSet <- readRDS("../../Result/Biomarker/Logistic_model/species_profile_Discovery.RDS")
ExprSet_Val <- readRDS("../../Result/Biomarker/Logistic_model/species_profile_Validation.RDS")

CRC_HC_PSS <- fread("../../Result/Biomarker/Logistic_model/CRC_HC_DEA_signature-2021-11-10.csv")
CRC_nonCRC_PSS <- fread("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_signature-2021-11-10.csv")
```


### Logistic Ridge regression model 
```{r}
build_Ridge <- function(dataset=ExprSet,
                        testres=CRC_HC_PSS,
                        group_name=subgrp[c(1,3)],
                        Prop=1, 
                        kfold=10,
                        splitPro=0.5){
  
  # dataset=ExprSet
  # testres=CRC_HC_PSS
  # group_name=subgrp[c(1,3)]
  # Prop=1
  # kfold=10
  # splitPro=0.5

  # Protein signatures
  dat_DEP <- testres %>% dplyr::select(Protein, Proportion) %>%
      filter(Proportion >= Prop)    

  # dataset of DEP
  if(length(group_name) == 3){
    group_name2 <- c("non_CRC", "CRC")
    pheno <- pData(dataset) %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[2], group_name2[2], group_name2[1])) %>%
      mutate(Group=factor(Group, levels = group_name2))
    group_name <- group_name2
  }else{
    pheno <- pData(dataset) %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name))    
  }
  edata <- exprs(dataset)[dat_DEP$Protein, rownames(pheno)] %>% data.frame()
  
  

  mdat_temp <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                         dplyr::select(SampleID, Group),
                       edata %>% t() %>% data.frame() %>%
                         rownames_to_column("SampleID"),
                       by = "SampleID") %>%
      column_to_rownames("SampleID")
    
  # Before building model, performing Zscore standardization
  mdat <- cbind(mdat_temp$Group, scale(mdat_temp %>% dplyr::select(-Group)) %>% data.frame())
  colnames(mdat)[1] <- "Group"
  
  # Data Partition
  #set.seed(123)
  require(sampling)
  sample <- sampling::strata(mdat, 
                     stratanames = "Group", 
                     size = round(as.numeric(table(mdat$Group)) * splitPro),
                     method = "srswor")
  trainData <- mdat[sample$ID_unit, ]
  testData <- mdat[-sample$ID_unit, ]
  print("The Group's number of TrainData")
  print(table(trainData$Group))
    
  dat_table <- as.matrix(trainData %>% dplyr::select(-Group))
  dat_target <- trainData$Group
  
  # preview data distribution
  # plot(density(dat_table))
  
  # determine folds
  if(dim(trainData)[1] > 10 * 8){
    kfold <- 10
  }else{
    kfold <- kfold
  }  
    
  # feature selection
  #set.seed(123)  
  cvfit <- cv.glmnet(x=dat_table,
                     y=dat_target,
                     family='binomial',
                     nfolds = kfold,
                     alpha = 0,
                     nlambda = 100,
                     type.measure = "auc")
  #plot(cvfit)
  predicted <- predict(cvfit, as.matrix(testData[, -1]), type = "response", s = cvfit$lambda.min)
    
  # AUC
  require(ggplot2)
  require(pROC)
  rocobj <- roc(testData$Group, predictor=predicted[, 1])
  
  # AUC 95% CI 
  auc_95CI <- ci.auc(rocobj, conf.level = 0.95)
  auc_95CI_value <- round(as.numeric(auc_95CI), 3)
  auc_95CI_label <- paste0("AUC = ", auc_95CI_value[2], " (", 
                            auc_95CI_value[1], ", ", 
                           auc_95CI_value[3], ")")

  auc <- round(auc(testData$Group, predictor=predicted[, 1]), 4)
  
  pl <- ggroc(rocobj, color = "red", linetype = 1, size = 1, alpha = 1, legacy.axes = T)+
                  geom_abline(intercept = 0, slope = 1, color="grey", size = 1, linetype=1)+
                labs(x = "False Positive Rate (1 - Specificity)",
                     y = "True Positive Rate (Sensivity or Recall)")+
                # annotate("text",x = .75, y = .25, label=paste("AUC =", auc),
                #          size = 5, family="serif")+
                annotate("text",x = .75, y = .25, label=auc_95CI_label,
                         size = 5, family="serif")+    
                coord_cartesian(xlim = c(0, 1), ylim = c(0, 1))+
                theme_bw()+
                theme(panel.background = element_rect(fill = 'transparent'),
                      axis.ticks.length = unit(0.4, "lines"), 
                      axis.ticks = element_line(color='black'),
                      axis.line = element_line(size=.5, colour = "black"),
                      axis.title = element_text(colour='black', size=12,face = "bold"),
                      axis.text = element_text(colour='black',size=10),
                      text = element_text(size=8, color="black", family="serif")) 
  
  
  # predictors
  predictors <- data.frame(as.matrix(coef(cvfit, cvfit$lambda.min))) %>%
          setNames("Score") %>%
          rownames_to_column("Protein") %>%
          slice(-1) %>%
          filter(Score!=0) %>%
      inner_join(testres, by = "Protein")

  res <- list(fit=cvfit,
              roc=rocobj,
              auc=auc,
              auc_CI=auc_95CI_label,
              pl=pl,
              marker=predictors)
  
  return(res)
}


model_validation <- function(dataset=ExprSet_Val,
                             datmarker=CRC_nonCRC_Ridge$marker$Protein,
                             datfit=CRC_nonCRC_Ridge$fit,
                             group_name=subgrp[1:3]){

  # dataset=ExprSet_Val
  # datmarker=CRC_nonCRC_Ridge$marker$Protein
  # datfit=CRC_nonCRC_Ridge$fit
  # group_name=subgrp[1:3]

  # dataset of DEP
  if(length(group_name) == 3){
    group_name2 <- c("non_CRC", "CRC")
    pheno <- pData(dataset) %>%
      filter(Group%in%group_name) %>%
      mutate(Group=ifelse(Group == group_name2[2], group_name2[2], group_name2[1])) %>%
      mutate(Group=factor(Group, levels = group_name2))
    group_name <- group_name2
  }else{
    pheno <- pData(dataset) %>%
      filter(Group%in%group_name) %>%
      mutate(Group=factor(Group, levels = group_name))
  }

  profile <- exprs(dataset)
  edata <- profile[, colnames(profile)%in%rownames(pheno)] %>%
    data.frame() %>%
    rownames_to_column("FeatureID") %>%
    filter(FeatureID %in% datmarker) %>%
    column_to_rownames("FeatureID")
    
  # Data of whole feature
  mdat_temp <- inner_join(pheno %>% rownames_to_column("SampleID") %>%
                       dplyr::select(SampleID, Group),
                     edata %>% t() %>% data.frame() %>%
                       rownames_to_column("SampleID"),
                     by = "SampleID") %>%
    column_to_rownames("SampleID")
  
  # Before building model, performing Zscore standardization
  mdat <- cbind(mdat_temp$Group, scale(mdat_temp %>% dplyr::select(-Group)) %>% data.frame())
  colnames(mdat)[1] <- "Group" 
  
  #datfit <- datmodel$fit

  # evaluate the performance of model
  predict_class <- predict(datfit, as.matrix(mdat[, -1]), type = "class")
  print(caret::confusionMatrix(factor(predict_class, levels = group_name), 
                               factor(mdat$Group, levels = group_name)))

  # if(nlevels(pred_raw$PredictedGroup) > 1){
  #   print(caret::confusionMatrix(pred_raw, mdat$Group))    
  # }
  pred_prob <- predict(datfit, as.matrix(mdat[, -1]), type = "response", s=datfit$lambda.min)

  # define roc and calculate AUC
  rocobj <- roc(mdat$Group, pred_prob[, 1])
  
  # AUC 95% CI 
  auc_95CI <- ci.auc(rocobj, conf.level = 0.95)
  auc_95CI_value <- round(as.numeric(auc_95CI), 3)
  auc_95CI_label <- paste0("AUC = ", auc_95CI_value[2], " (", 
                            auc_95CI_value[1], ", ", 
                           auc_95CI_value[3], ")")
  
  auc <- round(auc(mdat$Group, pred_prob[, 1]), 3)
  roc <- tibble(tpr=rocobj$sensitivities,
                fpr=1 - rocobj$specificities)
  pl <- ggplot(data=roc, aes(x=fpr, y=tpr))+
            geom_path(color="red", size=1)+
            geom_abline(intercept=0, slope=1, color="grey", size=1, linetype=2)+
            labs(x = "False Positive Rate (1 - Specificity)",
                 y = "True Positive Rate (Sensivity or Recall)")+
            # annotate("text", x=.75, y=.25, label=paste("AUC =", auc),
            #          size=5, family="serif")+
            annotate("text", x=.75, y=.25, label=auc_95CI_label,
                     size=5, family="serif")+    
            coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
            theme_bw()+
            theme(panel.background = element_rect(fill="transparent"),
                  axis.ticks.length = unit(0.4, "lines"),
                  axis.ticks = element_line(color="black"),
                  axis.line = element_line(size=.5, color="black"),
                  axis.title = element_text(color='black', size=12, face="bold"),
                  axis.text = element_text(color='black', size=10),
                  text = element_text(size=8, color="black", family="serif"))

  # Accuracy
  if(length(unique(predict_class)) > 1){
    CFM <- caret::confusionMatrix(factor(predict_class, levels = group_name), 
                                  factor(mdat$Group, levels = group_name))
    CFM_index <- data.frame(Value=CFM$byClass) %>%
      rownames_to_column("Index") %>%
      rbind(data.frame(Index="Accuracy", Value=as.numeric(CFM$overall[1])))
  }else{
    CFM <- NA
    CFM_index <- NA
  }

  res <- list(roc=rocobj,
              auc=auc,
              auc_CI=auc_95CI_label,
              CFM=CFM,
              index=CFM_index,
              pl=pl)

  return(res)
}


model_performance <- function(rocobject=CRC_nonCRC_Ridge$roc,
                              auc_CI=CRC_nonCRC_Ridge$auc_CI,
                              tag="CRC vs Non-CRC"){
  
  # rocobject=CRC_nonCRC_Ridge$roc
  # auc_CI=CRC_nonCRC_Ridge$auc_CI
  # tag="CRC vs Non-CRC"
  
  require(pROC)
  # 95% CI 
  CI95_Normal <- function(x, tag){
    # Calculating a Confidence Interval From a Normal Distribution
    x_length <- length(x)
    x_mean <- signif(mean(x), 3)
    x_sd <- sd(x)
    error <- qnorm(0.975) * x_sd/sqrt(x_length)
    left <- signif(x_mean - error, 3)
    right <- signif(x_mean + error, 3)
    res <- paste0(x_mean, " (", left, ", ", right, ")")
    #res <- paste0(tag, " = ", x_mean, " (", left, ", ", right, ")")
    return(res)
  }
  
  # Sensitivity and Specificity and positive predictive value
  ss <- coords(rocobject, "all", ret="sensitivity", transpose = TRUE)
  ss_95CI <- CI95_Normal(ss, "sensitivity")

  sf <- coords(rocobject, "all", ret="specificity", transpose = TRUE)
  sf_95CI <- CI95_Normal(sf, "specificity")

  PPV <- coords(rocobject, "all", ret="precision", transpose = TRUE)
  PPV_95CI <- CI95_Normal(PPV[!is.nan(PPV)], "precision")
  
  acc <- coords(rocobject, "all", ret="accuracy", transpose = TRUE)
  acc_95CI <- CI95_Normal(acc, "accuracy")  
  
  # AUC
  auc_95CI <- gsub("AUC = ", "", auc_CI) 
  
  res <- data.frame(Name=tag, AUC=auc_95CI, Accuracy=acc_95CI,
                    Sensitivity=ss_95CI, Specificity=sf_95CI, Precision=PPV_95CI)
  
  return(res)

}

if(!dir.exists("../../Result/Biomarker/Logistic_model")){
  dir.create("../../Result/Biomarker/Logistic_model", recursive = T)
}
```


### CRC vs HC 
```{r, fig.width=12, fig.height=5}
CRC_HC_Ridge <- build_Ridge(dataset=ExprSet,
                            testres=CRC_HC_PSS,
                            group_name=subgrp[c(1,3)],
                            Prop=0.7, 
                            kfold=10,
                            splitPro=0.5)

model_performance(rocobject=CRC_HC_Ridge$roc,
                  auc_CI=CRC_HC_Ridge$auc_CI,
                  tag="HC vs CRC")

CRC_HC_Val <- model_validation(dataset=ExprSet_Val,
                               datmarker=CRC_HC_Ridge$marker$Protein,
                               datfit=CRC_HC_Ridge$fit,
                               group_name=subgrp[c(1,3)])
CRC_HC_Val$pl

CRC_HC_filename_Ridge <- paste0("../../Result/Biomarker/Logistic_model/CRC_HC_DEA_Ridge-", Sys.Date(), ".RDS")
saveRDS(CRC_HC_Ridge, file = CRC_HC_filename_Ridge, compress = TRUE)
CRC_HC_filename_Ridge_val <- paste0("../../Result/Biomarker/Logistic_model/CRC_HC_DEA_Ridge_Validation-", Sys.Date(), ".RDS")
saveRDS(CRC_HC_Val, file = CRC_HC_filename_Ridge_val, compress = TRUE)
```


### CRC vs non_CRC 
```{r, fig.width=12, fig.height=5}
CRC_nonCRC_Ridge <- build_Ridge(dataset=ExprSet,
                            testres=CRC_nonCRC_PSS,
                            group_name=subgrp[c(1:3)],
                            Prop=.7, 
                            kfold=10,
                            splitPro=0.5)

model_performance(rocobject=CRC_nonCRC_Ridge$roc,
                  auc_CI=CRC_nonCRC_Ridge$auc_CI,
                  tag="nonCRC vs CRC")

CRC_nonCRC_Val <- model_validation(dataset=ExprSet_Val,
                                   datmarker=CRC_nonCRC_Ridge$marker$Protein,
                                   datfit=CRC_nonCRC_Ridge$fit,
                                   group_name=subgrp[c(1:3)])

CRC_nonCRC_filename_Ridge <- paste0("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_Ridge-", Sys.Date(), ".RDS")
saveRDS(CRC_nonCRC_Ridge, file = CRC_nonCRC_filename_Ridge, compress = TRUE)
CRC_nonCRC_filename_Ridge_val <- paste0("../../Result/Biomarker/Logistic_model/CRC_nonCRC_DEA_Ridge_Validation-", Sys.Date(), ".RDS")
saveRDS(CRC_nonCRC_Val, file = CRC_nonCRC_filename_Ridge_val, compress = TRUE)
```


### systemic information
```{r}
sessionInfo()
```

